Lessons learned and paths forward for mass dog vaccination in Madagascar: a case study of pilot canine rabies vaccination campaigns in Moramanga District

### 2021-02-04 manuscript.Rmd
adding author list & acknowledgements

---
title: "Lessons learned and paths forward for mass dog vaccination in Madagascar: a case study of pilot canine rabies vaccination campaigns in Moramanga District"
author: "Malavika Rajeev"
date: "11/19/2020"
output:
  word_document: default
  html_document: default
bibliography: format/references.bib
---

@@

@@

Canine rabies results in an estimated 60,000 human deaths per year globally. These deaths are entirely preventable: prompt post-exposure prophylaxis (PEP) of humans is highly effective at preventing death and mass dog vaccination can interrupt transmission in domestic dogs and eventually lead to disease elimination. Annual vaccination campaigns achieving atleast 70% coverage are the recommended target for controlling rabies in domestic dog populations. However, achieving this coverage target in low and middle income countries where the virus is endemic can be challenging due to economic, ecological, sociocultural, and political reasons.

@@

The Moramanga District is located midway between the central highlands and the east coast of Madagascar, at an average altitude of 936 m. It comprises 21 communes and NN fokontany (sub-commune), covering approximately 7150 km^2^ with a human population of x (2018 census prelim data). Previous work in the district has established a high burden of rabies exposures and deaths despite the availability of post-exposure prophylaxis at the district hospital. While Moramanga is relatively close to the capital city of Antananarivo (~ 3 hrs by bus), within the district, travel times between locations are highly variable, with much of the population living in more rural areas with limited access to roads and transportation [cite my other paper]. Before 2018, there were limited animal rabies vaccination services, with most vaccines available in urban communes and owners charged > 15,000 Ar (~ x USD) per vaccine administered.

@@

#### Data and ethics statement
All data were analysed in `r R.version$version.string` [@R2020], largely using the `tidyverse` package suite [@tidyverse]. Geospatial data were mapped using the sf [@sf] package. All data and code are available at [https://github.com/mrajeev08/mora_vax](https://github.com/mrajeev08/mora_vax). The vaccinations were part of a public health campaign and routine veterinary service provisioning carried out by the local veterinary officals, the NGOs involved, and in partnership with the Ministry of Public Health and the Department of Veterinary Services at the national level. The Madagascar Dog Initiative also maintains institutional approvals (IACUC # at Institution X) and national research permits (MICET permit #) for it's research and volunteer programs.

@@

While less than 20% of owners in Moramanga Ville reported that their animals free roaming, fenced animals were frequently observed outside fences during the campaign, and thus the majority of animals could be classified as semi-confined in the more urban township of Moramanga Ville, and free-roaming in the rural setting of Andasibe (approximately 70% of owners reported their dogs as free-roaming, Table 1). In 2019, the DV also asked owners how far they travelled approximately in meters, and `r dist_over_100`% of people travelled less than 1 km to reach the vaccination point.

@@

```{r fig3}
knitr::include_graphics("figs/fig3.jpeg")
```
#### Figure 3. Dog demography and implications for campaign strategies
*A) The age pyramid for vaccinated dogs by sex. B) Bootstrapped estimates of the proportion of the population in each age class (dark blue) compared to estimates from the demographic models fit to these age data. C) Parameter estimates for annual fertility rates and adult survival probability, with estimates in orange showing the parameter estimates which result in positive population growth. B) Simulated vaccination coverage (N = 1000) using the demographic parameters from (C) in a hypothetical commune with 1000 dogs for three different campaign strategies: 1) annual vaccination campaigns targeting dogs of all ages (purple), 2) Routine vaccination of puppies at 3 months of age, and 3) a combined strategy with campaigns annually and continuous puppy vaccination in between campaigns.*
## Discussion

Main findings para (feasible, should not charge, costs vs. coverage).

@@

There were several limitations to our analyses. Owner reports of willingness to pay, age of animals, and distance travelled to the vaccination point likely all suffer from recall bias and uncertainty. However, our findings from these data are consistent with previous literature. We estimate that the dog population is male-biased and skewed towards puppies, but that pups likely suffer high mortatlity in the first year of life, consistent with studies from other sub-Saharan African countries. Other studies have also shown that implementing fees for vaccination greatly reduces the efficacy of campaigns. Similarly, distance to campaign points has also been identified as major barrier to vaccination, and as with other studies we found that the majority of people report traveling less than 1km to reach a vaccination point. Dog population estimates were also largely lacking, but we were able to ground our estimates of human-to-dog ratios from a recent household survey study in the distict and used a range that is consistent with the literature. *Estimates between the two methods were comparable.* We also make several simplifying assumptions in our model of vaccination strategies: we assume that immunity wanes after three years (simulating vaccination with a long-lasting vaccine such as Nobivac, compared to the 1 year efficacy estimated for the Rabisin vaccine); we assume that vaccinated dogs that survive from year to year are revaccinated in subsequent campaigns; and in our age-structured models we do not account for population carrying capacity, but rather simulate over a short time frame over which growth estimates remain reasonable.

*Costs for the campaign did not include volunteer costs. Point to volunteer issues.*  Cost not included was PEP for exposures during a campaign (mostly cat scratches). Getting vaccinators prep. PrEP and exposures during vax, we had N exposures. Puppies were the easiest to handle for both owners and vaccinators and vax is effective in them. Better dosing guidelines. Longer lasting vax. *We weren't that good person per day*

One key aspect we do not consider is the feedback loops between vaccination and demography.  Estimates of the effects of vaccination on dog demography are mixed, but vaccination could potentially increase survival of dogs. If dog population growth is driven by survival, then this could mean that increased vaccination results in increased population growth. However, if growth is driven more by demand from human communities, then vaccination could stabilize the population and reduce turnover of the population. Improving dog population management, encouraging responsible pet ownership practices, and increasing veterinary services could all complement vaccination efforts. *Overall high turnover of the dog pop makes it harder to.*

In these settings with high dog ownership, moving towards community-based vaccination strategies, could be a highly effective way to achieve sufficient and consistent coverage, particulalry in rural areas where static point campaigns may be less likely to reach. During our campaign, we found that puppies at approximately 3 months old were easy to handle for both vaccinators and owners. Puppy vaccination could therefore use local officials emmbedded in communities (along the lines of community health workers) to implement vaccination, particularly given recent findings on thermotolerance of rabies vaccines and locally manufactured methods for maintaing temperatures required for sustained vaccine storage (up to 3 months). *Covid? Accessibility?*

Paying people is important hard/relieving pressures and making sure it's viewed as a public good (Ahmed's paper). Veterinary students?
Ultimately push for govt investment and viewing as a public good. The veterinary workforce (relieving pressures on them, challenges with private-public roles).. Modeling after childhood vax = routine through local offices of puppies + supplemental activities by trained vets annually to catch unvax (feral, oral).

Policy recommendations:
- Dog vax as a public good, remove barriers for dog owners & incentivize vets/etc/ remove economic pressures
- Community based vax delivery particulalry in rural areas (could lower cost while improving coverage)
- Easier ways to estimate coverage (& coupling with surveillance to assess impact)

## Acknowledgements

We thank vets livestock officers volunteers fokontany leaders government officials. We want to especially acknowledge the important contributions made by Annie Yang, an undergraduate researcher and volunteer with Princeton University and the Madagascar Dog Initiative who tragically passed away in a car accident before this research was published.

>>>>

---
title: "Lessons learned and paths forward for mass dog vaccination in Madagascar: a case study of pilot canine rabies vaccination campaigns in Moramanga District"
author:
  - Caitlynn Filla:
      insitute: [florida]
      equal_contributor: "yes"
  - Malavika Rajeev:
      email: mrajeev@princeton.edu
      institute: [princeton]
      correspondence: true
      equal_contributor: "yes"
  - Zoavina Randriana:
      institute: [maddog]
  - Chantal Hanitriniana:
      institute: [uoa]
  - Radoniaina R. Rafaliarison:
      institute: [maddog]
  - ?Glenn Edosoa:
      institute: [oms]
  - ?Mamitiana Andriamananjara:
      institute: [dsv]
  - Fenomanana Daniel:
      institute: [maddog]
  - Tara A. Clarke:
      institute: [maddog]
  - Zachary Farris:
      institute: [maddog, appstate]
  - Terry Stone:
      institute: [tad]
  - Jochem Lastdrager:
      institute: [tad]
  - Tsiky Rajaonarivelo:
      institute: [maddog]
  - Katie Hampson:
      institute: [glasgow]
  - C. Jessica E. Metcalf:      
      institute: [princeton]
  - Kim Valenta:
      institute: [maddog, florida]
institute:
  - princeton: Department of Ecology and Evolutionary Biology, Princeton University, Princeton, United States
  - oms: Chargé des Maladies Tropicales Négligées, Organisation mondiale de la Santé Madagascar, Antananarivo, Madagascar
  - florida: University of Florida, Department of Anthropology, Gainesville, Florida, USA
  - maddog: The Mad Dog Initiative, Akanin’ny Veterinera Akaikiniarivo, Ambatobe, Antananarivo 101, Madagascar
  - tad: Travelling Animal Doctors, 919 Crossan Rd, Newark, Delaware, USA 
  - uoa: Mention Zoologie et Biodiversité Animale, Faculté des Sciences, Université d’Antananarivo, Antananarivo, Madagascar
  - glasgow: Boyd Orr Centre for Population and Ecosystem Health, Institute of Biodiversity, Animal Health and Comparative Medicine, University of Glasgow, Glasgow, UK
  - appstate: Appalachian State University, Department of Health and Exercise Science, Boone, North Carolina, USA
  - dsv: Direction des Services Vétérinaires, Ministère chargé de l’Agriculture et de l’Élevage, Antananarivo, Madagascar
output:
  bookdown::word_document2:
    pandoc_args:
      - --lua-filter=format/scholarly-metadata.lua
      - --lua-filter=format/author-info-blocks.lua
      - --lua-filter=format/pagebreak.lua
  html_document: default
bibliography: format/references.bib
---

@@

Canine rabies causes an estimated 60,000 human deaths per year, but these deaths are preventable through post-exposure prophylaxis of people and vaccination of domestic dogs. Dog vaccination campaigns acheiving coverage of 70% at the population level are effective at interrupting transmission and eventually can lead to elimination of the disease regionally. Given this evidence, the WHO and it's partners have set a global goal to eliminate deaths due to canine rabies by the year 2030. Here, we report on initial lessons learned during pilot dog vaccination campaigns in the Moramanga District of Madagascar, a country with a high estimated burden of rabies and limited control in the dog population. We compare two different vaccination strategies: a larger volunteer driven effort to vaccinate dogs in two communes using static point vaccination, and continuous vaccination as part of routine veterinary services in the district by the District Veterinarian. We compared costs, time, personnel, and coverage acheived by each of the campaigns. We also use age data from the campaigns to estimate key demographic parameters, and use these estimates to simulate different vaccination strategies. Overall we found that dog vaccination was feasible and that most dogs were accessible to vaccination. We find that there are tradeoffs between the vaccination strategies, with the first campaign acheiving higher coverage, but with more resources and with more limited geographic coverage compared to the second campaign with continuous delivery. Our modeling results suggest that   that targeting puppies through community based vaccination efforts could hypothetically acheive coverage levels sufficient to interrupt transmission. Moving forward, our results show that mass dog vaccination is feasible and can acheive high coverage in Madagascar, however innovative strategies and an investment in dog vaccination as a public good will be required to maintain the high and sustained levels of coverage needed for disease elimination.

@@

Canine rabies results in an estimated 60,000 human deaths per year globally. These deaths are entirely preventable: prompt post-exposure prophylaxis (PEP) of humans is highly effective at preventing death and mass dog vaccination can interrupt transmission in domestic dogs and eventually lead to disease elimination. The WHO and its partners have set a goal to eliminate human deaths due to canine rabies by the year 2030 ('ZeroBy30'). Annual vaccination campaigns achieving atleast 70% coverage are the recommended target for controlling rabies in the domestic dog populations. However, achieving this coverage target in low and middle income countries where the virus is endemic can be challenging due to economic, ecological, sociocultural, and political reasons.

@@

The Moramanga District is located midway between the central highlands and the east coast of Madagascar, at an average altitude of 936 m. It comprises 21 communes and NN fokontany (sub-commune), covering approximately 7150 km^2^ with a human population of x (2018 census prelim data). Urban center of Moramanga (cite Rila's paper). Previous work in the district has established a high burden of rabies exposures and deaths despite the availability of post-exposure prophylaxis at the district hospital. While Moramanga is relatively close to the capital city of Antananarivo (~ 3 hrs by bus), within the district, travel times between locations are highly variable, with much of the population living in more rural areas with limited access to roads and transportation [cite my other paper]. Before 2018, there were limited animal rabies vaccination services, with most vaccines available in urban communes and owners charged > 15,000 Ar (~ x USD) per vaccine administered.

@@

#### Data and ethics statement
All data were analysed in `r R.version$version.string` [@R2020], largely using the `tidyverse` package suite [@tidyverse]. Geospatial data were mapped using the sf [@sf] package. All data and code are available at [https://github.com/mrajeev08/mora_vax](https://github.com/mrajeev08/mora_vax). The vaccinations were part of a public health campaign and routine veterinary service provisioning carried out by the local veterinary officals, the NGOs involved, and in partnership with the Ministry of Public Health and the Department of Veterinary Services at the national level. The Madagascar Dog Initiative also maintains institutional approvals (IACUC # at Institution X) and national research permits (MICET permit #) for it's research and volunteer programs [Kim to fill in].

@@

While less than 20% of owners in Moramanga Ville reported that their animals were free roaming, fenced animals were frequently observed outside fences during the campaign, and thus the majority of animals could be classified as semi-confined in the more urban township of Moramanga Ville, and free-roaming in the rural setting of Andasibe (approximately 70% of owners reported their dogs as free-roaming, Table 1). In 2019, the DV also asked owners how far they travelled approximately in meters, and `r dist_over_100`% of people travelled less than 1 km to reach the vaccination point.

@@

```{r fig3}
knitr::include_graphics("figs/fig3.jpeg")
```
#### Figure 3. Dog demography and implications for campaign strategies
*A) The age pyramid for vaccinated dogs by sex. B) Bootstrapped estimates of the proportion of the population in each age class (dark blue) compared to estimates from the demographic models fit to these age data. C) Parameter estimates for annual fertility rates and adult survival probability, with estimates in orange showing the parameter estimates which result in positive population growth. B) Simulated vaccination coverage (N = 1000) using the demographic parameters from (C) in a hypothetical commune with 1000 dogs for three different campaign strategies: 1) annual vaccination campaigns targeting dogs of all ages (purple), 2) Routine vaccination of puppies at 3 months of age, and 3) a combined strategy with campaigns annually and continuous puppy vaccination in between campaigns.*

## Discussion

@@

There were several limitations to our analyses. Owner reports of willingness to pay, age of animals, and distance travelled to the vaccination point likely all suffer from recall bias and uncertainty. However, our findings from these data are consistent with previous literature. We estimate that the dog population is male-biased and skewed towards puppies, but that pups likely suffer high mortatlity in the first year of life, consistent with studies from other sub-Saharan African countries. Other studies have also shown that implementing fees for vaccination greatly reduces the efficacy of campaigns. Similarly, distance to campaign points has also been identified as major barrier to vaccination, and as with other studies we found that the majority of people report traveling less than 1km to reach a vaccination point. Dog population estimates were also largely lacking, but we were able to ground our estimates of human-to-dog ratios from a recent household survey study in the distict and used a range that is consistent with the literature, and coverage estimates using HDRs were comparable to estimates from post-vaccination transects. We also make several simplifying assumptions in our model of vaccination strategies: we assume that immunity wanes after three years (simulating vaccination with a long-lasting vaccine such as Nobivac, compared to the 1 year efficacy estimated for the Rabisin vaccine); we assume that vaccinated dogs that survive from year to year are revaccinated in subsequent campaigns; and in our age-structured models we do not account for population carrying capacity, but rather simulate over a short time frame over which growth estimates remain reasonable.

Our estimates of costs per animal vaccinated are in line with recent estimates from other settings, but on the lower end. In our estimates, we excluded costs associated with international volunteers (i.e. airfare and visa costs) as well as costs of transportation, as these were donated to the campaign. We also did not include estimates of pre-exposure prophylaxis as all of our volunteers and vaccinators had received the course prior to the campaigns. In a rabies endemic situation, it will be necessary to have both PrEP and PEP readily available for vaccinators. While the volunteer-led effort resulted in significant financial and personnel resources being devoted to the campaign, costs were higher per animal and overall. In addition, it can be difficult for NGO and volunteer based efforts to sustain these types of efforts given unpredictable funding, time commitments, and turnover in staff. The routine vaccination strategy through the DV was less costly, with the majority of the costs directly related to vaccination. Transport to locations was not included in these estimates either, as the DV had their own vehicle and largely delivered the vaccinations as part of her routine services.

One key aspect we do not consider is the feedback loops between vaccination and demography.  Estimates of the effects of vaccination on dog demography are mixed, but vaccination could potentially increase survival of dogs. If dog population growth is driven by survival, then this could mean that increased vaccination results in increased population growth. However, if growth is driven more by demand from human communities, then vaccination could stabilize the population and reduce turnover of the population. Improving dog population management, encouraging responsible pet ownership practices, and increasing veterinary services could all complement vaccination efforts. Overall, high population turnover of the dog population erodes coverage, therefore strategies to stabilize the population would be beneficial to improving vaccination coverage.

In these settings with high dog ownership, moving towards community-based vaccination strategies, could be a highly effective way to achieve sufficient and consistent coverage, particulalry in rural areas where static point campaigns may be less likely to reach. During our campaign, we found that puppies at approximately 3 months old were easy to handle for both vaccinators and owners. Puppy vaccination could therefore use local officials emmbedded in communities (along the lines of community health workers) to implement vaccination, particularly given recent findings on thermotolerance of rabies vaccines and locally manufactured methods for maintaing temperatures required for sustained vaccine storage (up to 3 months). Incentivizing vaccinators appropriately will be a key challenge, as currently providing no-cost rabies vaccination is not seen as part of routine duties for district veterinarians or for livestock officers, and these officials often required additional per diems. Expanding veterinary services across the country and releiving financial pressures on veterinarians and animal health workers could greatly mitigate these issues in Madagascar. For volunteer based efforts, veterinary students could be a local volunteer pool, however similarly to international volunteers, they require food, transport, and lodging during vaccination campaigns when not based in the communities where they study or live, adding to overall costs.

Overall, our results suggest that dog vaccination is a feasible and feasible strategy for controlling canine rabies. It will be key to push rabies vaccination as a public good and to remove barriers for dog owners as well as to incentivize veterinarians and other animal health workers to implement vaccination. Borrowing strategies from human vaccination efforts, i.e. childhood vaccination and using community health workers to provide routine services, could be a way to improve vaccination coverage and reduce costs, particularly in rural communities. To monitor the success of these campaigns, it will be critical to develop efficient and effective methods to estimate vaccination coverage and the impact of campaigns on rabies incidence through robust surveillance systems. Given it's island context, Madagascar could be a model for rabies elimination, with limited chances for reintroduction from outside the island, and strategies that tackle the context specific challenges could be a path forward for the country to reach ZeroBy30.

## Acknowledgements
We thank all the veterinarians, livestock officers, local officials and leaders, NGO staff, and volunteers involved in each of the campaigns. We are grateful to staff and officials at the Department of Veterinary Services, the Ministry of Public Health, and the Institut Pasteur de Madagascar for their support and technical assistance. In particular, we thank Angelique Ferreira, Jean Hyacinthe Randrianarisoa, Ranaivoarimanana, Fierenantsoa Randriamahatana, Esther Noiarisaona, Cara Brook, and Christian Ranaivoson, Rila Ratovoson, and Claire LeBlanc. The campaign costs were supported by donations from Lush Cosmetics and [Kim to insert]. Parts of this work were funded by grants from the Center for Health and Wellbeing and the Department of Ecology and Evolutionary Biology at Princeton University to MR. MR was supported by an NSF Graduate Research Fellowship and a Princeton Insitute for Regional and International Studies Graduate Fellowship. KH is supported by the Wellcome Trust (207569/Z/17/Z).

We want to particularly acknowledge the contributions to this work by Annie Yang, an undergraduate researcher and volunteer with Princeton University and the Madagascar Dog Initiative who tragically passed away in a car accident before this research was published.
